Solving Graph Homomorphism and Subgraph Isomorphism Problems Faster Through Clique Neighbourhood Constraints

### 2021-05-20 paper/paper.tex
Revert "Completely anonymous"
This reverts commit 7b978a5c313d109c7c589c4e01ccc510dcc9999f.
Tweaks

\author{Paper 4726}

@@

As well as questions over exactly how edges are represented and mapped, some applications would
prefer not to specify injectivity. A \emph{graph homomorphism} is a function between two graphs that
maps adjacent vertices to adjacent vertices, with no injectivity requirement, whilst a \emph{locally
injective homomorphism} is one which is injective when restricted to any individual vertex and its
neighbourhood. Both are better models for some applications
\cite{DBLP:journals/dam/FialaKK01,DBLP:conf/semweb/Baget05,DBLP:conf/iccs/CorbyF07,DBLP:journals/csr/FialaK08,DBLP:journals/pvldb/FanLMWW10,DBLP:journals/tcs/ChaplickFHPT15}.
(There is also an edge-surjective version of the problem known as an \emph{epimorphism}
\cite{DBLP:journals/dam/GayFMSS14}.)

@@

We begin by introducing notation and terminology, and by giving the theory which
support our implementation.

@@

\begin{proposition}[paths are preserved by subgraph isomorphisms]\label{proposition:distances}For
    the problem of finding a subgraph isomorphism $i$ from a graph $G$ to a graph $H$, the following
    constraints are implied for any pair of vertices $v, w \in \vertexset(G)$:
    \begin{enumerate}
        \item The distance between $v$ and $w$ is at least the distance between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/AudemardLMGP14}.
        \item If there are at least $k$ simple
            paths of length exactly $\ell$ between $v$ and $w$, then there must be at least $k$
            simple paths of length exactly $\ell$ between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/McCreeshP15}.
    \end{enumerate}
\end{proposition}

It is easy to verify that the first of these two properties is valid for \emph{any} homomorphism.
The second property does not hold for homomorphisms in general: for example, a pair of vertices
connected by three paths of length two may be mapped onto a pair of vertices connected by a single
path of length two (see \cref{figure:counterexamples}, top right). The second property also does not
hold in full generality for local injectivity, because, for example, a path of length three may be
mapped onto a triangle (\cref{figure:counterexamples}, bottom left): while standard injectivity
prevents any kind of merging, local injectivity only prevents two vertices from being merged if they
are both in the neighbourhood of the same vertex. However, we observe that a weaker result does
hold:

@@

Rather than using distance properties directly as \citet{DBLP:conf/cp/AudemardLMGP14} did,
\citet{DBLP:conf/cp/McCreeshP15} introduced the notion of \emph{supplemental} graphs, with the idea
that a constraint programming algorithm can search for a mapping which is simultaneously a subgraph
isomorphism between several different pairs of graphs. Let $G^d$ be the graph with the same set of
vertices as $G$, but an edge between vertices $v$ and $w$ if the distance between $v$ and $w$ in $G$
is at most $d$.  Similarly, let $G^{n,\ell}$ be the graph with the same set of vertices as $G$, but
an edge between vertices $v$ and $w$ if there are at least $n$ simple paths of length exactly $\ell$
between $v$ and $w$ in $G$. Following on from \cref{proposition:distances,proposition:paths}, we
generalise the result of \citeauthor{DBLP:conf/cp/McCreeshP15} as follows.

@@

Given the complexity of modern solvers, a critical question is how we can be sure
that they are producing correct answers. The Glasgow Subgraph Solver's subgraph isomorphism and
clique algorithms are both \emph{certifying} \cite{DBLP:journals/csr/McConnellMNS11}: that is, they
can output a mathematical proof that they have reached a correct answer by sound reasoning, and this
proof log can be verified by the VeriPB proof checking tool
\cite{DBLP:conf/cp/GochtMMNPT20,DBLP:conf/ijcai/GochtMN20}.

Implementing proof-logging for the changes not involving clique-finding was routine---although proof
logging did catch an implementation bug where we were incorrectly applying degree-based filtering on
the $G^{2,n}$ graphs for locally injective homomorphisms, in violation of the counter-example to
\cref{corollary:lishapes}. For clique-based filtering, the process involved substantially more
technical work. Briefly, for a given domain vertex $v$ that cannot be mapped to a codomain vertex
$w$, it is possible to instruct the proof verifier to derive a set of conditional clique constraints
saying that ``either $v$ cannot be mapped to $w$, or the following clique-like problem must be
satisfiable'' by following steps similar to the maximum common subgraph to clique reduction
presented by \citet{DBLP:conf/cp/GochtMMNPT20}, and then to reuse clique
proof-logging techniques to show that this clique-like problem is unsatisfiable.

This situation is not entirely ideal. The proofs produced this way are inherently dependent upon a
\emph{particular} choice of domain and codomain vertices. When not proof logging, we only need to
solve a single maximum clique problem for each vertex in the domain, and a restricted clique problem
for each vertex in the codomain. With proof logging, we must instead solve restricted clique
problems for each assignment that is to be filtered, which is potentially a linear factor blowup.
This seems wasteful, since for any given codomain vertex these proofs are effectively the same up to
substitution of variable names. It would therefore seem desirable to extend the proof system with a
substitution rule which would allow such proofs to be recycled to avoid duplication.

@@

We now evaluate this approach empirically. Our experiments are performed on a cluster of machines
with dual Intel Xeon E5-2697A v4 CPUs with 512GBytes of RAM, running Ubuntu 18.04, and using a
timeout of one hour. Because injectivity relaxations have received less attention until now, for all
three problem variants we will use a suite of 14,621 benchmark instances designed for the subgraph
isomorphism problem \cite{DBLP:conf/lion/KotthoffMS16}. Note that 8,904 of these instances contain
at least one loop in the codomain graph, and so are trivial for the homomorphism problem---we still
include these to avoid using different y-axes on different plots.

@@

>>>>

\author{
Sonja Kraiczy$^1$
\and
Ciaran McCreesh$^2$
\affiliations
$^1$University of Oxford\\
$^2$University of Glasgow\\
\emails
ciaran.mccreesh@glasgow.ac.uk
}

@@

As well as questions over exactly how edges are represented and mapped, some applications would
prefer not to specify injectivity. A \emph{graph homomorphism} is a function between two graphs that
maps adjacent vertices to adjacent vertices, with no injectivity requirement, whilst a \emph{locally
injective homomorphism} is one which is injective when restricted to any individual vertex and its
neighbourhood. Both are better models for some applications
\cite{DBLP:journals/dam/FialaKK01,DBLP:conf/semweb/Baget05,DBLP:conf/iccs/CorbyF07,DBLP:journals/csr/FialaK08,DBLP:journals/pvldb/FanLMWW10,DBLP:journals/tcs/ChaplickFHPT15}.

@@

We begin by introducing notation and terminology, and by giving the theory which
supports our implementation.

@@

\begin{proposition}[paths are preserved by subgraph isomorphisms]\label{proposition:distances}For
    the problem of finding a subgraph isomorphism $i$ from a graph $G$ to a graph $H$, the following
    constraints are implied for any pair of vertices $v, w \in \vertexset(G)$:
    \begin{enumerate}
        \item The distance between $v$ and $w$ is at most the distance between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/AudemardLMGP14}.
        \item If there are at least $k$ simple
            paths of length exactly $\ell$ between $v$ and $w$, then there must be at least $k$
            simple paths of length exactly $\ell$ between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/McCreeshP15}.
    \end{enumerate}
\end{proposition}

It is easy to verify that the first of these two properties is valid for \emph{any} homomorphism.
The second property does not hold for homomorphisms in general: for example, a pair of vertices
connected by three paths of length two may be mapped onto a pair of vertices connected by a single
path of length two (see \cref{figure:counterexamples}, top right). The second property also does not
hold in full generality for local injectivity, because, for example, a path of length three may be
mapped onto a triangle (\cref{figure:counterexamples}, bottom left): while standard injectivity
prevents any kind of merging, local injectivity only prevents two vertices from being merged if they
are both in the neighbourhood of the same vertex. However, a weaker result does hold:

@@

Rather than using distance properties directly as \citet{DBLP:conf/cp/AudemardLMGP14} did,
\citet{DBLP:conf/cp/McCreeshP15} introduced the notion of \emph{supplemental} graphs, with the idea
that a constraint programming algorithm can search for a mapping which is simultaneously a subgraph
isomorphism between several different pairs of graphs. Let $G^d$ be the graph with the same set of
vertices as $G$, but with an edge between vertices $v$ and $w$ if the distance between $v$ and $w$
in $G$ is at most $d$.  Similarly, let $G^{n,\ell}$ be the graph with the same set of vertices as
$G$, but an edge between vertices $v$ and $w$ if there are at least $n$ simple paths of length
exactly $\ell$ between $v$ and $w$ in $G$. Following on from
\cref{proposition:distances,proposition:paths}, we generalise the result of
\citeauthor{DBLP:conf/cp/McCreeshP15} as follows.

@@

Given the complexity of modern solvers, a critical question is how we can be sure that they are
producing correct answers. The Glasgow Subgraph Solver's subgraph isomorphism and clique algorithms
are both \emph{certifying} \cite{DBLP:journals/csr/McConnellMNS11}: that is, they can output a
mathematical proof that they have reached a correct answer by sound reasoning, and this proof log
can be verified by the VeriPB proof checking tool
\cite{DBLP:conf/cp/GochtMMNPT20,DBLP:conf/ijcai/GochtMN20}. We therefore extended the solver's
existing proof logging support to cover our additions. For the changes not involving clique-finding,
this was entirely routine---although proof logging did catch an implementation bug where we were
incorrectly applying degree-based filtering on the $G^{2,n}$ graphs for locally injective
homomorphisms, in violation of the counter-example to \cref{corollary:lishapes}. For clique-based
filtering, the process involved substantially more technical work. Briefly, for a given domain
vertex $v$ that cannot be mapped to a codomain vertex $w$, it is possible to instruct the proof
verifier to derive a set of conditional clique constraints saying that ``either $v$ cannot be mapped
to $w$, or the following clique-like problem must be satisfiable'' by following steps similar to the
maximum common subgraph to clique reduction presented by \citet{DBLP:conf/cp/GochtMMNPT20}, and then
to reuse clique proof-logging techniques to show that this clique-like problem is unsatisfiable.  We
were thus able to produce and verifying proofs of correctness for medium-sized instances, giving us
confidence our implementation was correct.

This situation is not entirely ideal. The proofs produced this way are inherently dependent upon a
\emph{particular} choice of domain and codomain vertices. Thus, when proof logging, we must solve
restricted clique problems for each assignment that is to be filtered using a clique rule, rather
than for each clique found; this is potentially a linear factor blowup.  This seems wasteful, since
for any given codomain vertex these proofs are effectively the same up to substitution of variable
names. It would therefore seem desirable to extend the proof system with a substitution rule which
would allow such proofs to be recycled to avoid duplication.

@@

We now evaluate this approach empirically. Our experiments are performed on a cluster of machines
with dual Intel Xeon E5-2697A v4 CPUs with 512GBytes of RAM, running Ubuntu 18.04, and using a
timeout of one hour; we do not enable proof logging for these benchmarks since proof logging
introduces large slowdowns due to I/O costs \cite{DBLP:conf/cp/GochtMMNPT20}. Because injectivity
relaxations have received less attention until now, for all three problem variants we will use a
suite of 14,621 benchmark instances designed for the subgraph isomorphism problem
\cite{DBLP:conf/lion/KotthoffMS16}. Note that 8,904 of these instances contain at least one loop in
the codomain graph, and so are trivial for the homomorphism problem---we still include these to
avoid using different y-axes on different plots.

@@

\section*{Acknowledgements}

This work was supported by the Engineering and Physical Sciences Research Council [grant number
EP/P026842/1].
